---
title: "1. Introduction: We want cell topic estimation ASAP"
author: "Yongjin Park"
---

```{r include = FALSE}
library(ggplot2)
library(reshape2)
theme_set(theme_classic())
```

```{r simulation}
library(asapR)
library(pheatmap)
set.seed(1331)
d <- 500
n <- 5000
.rnorm <- function(d1,d2) matrix(rnorm(d1 * d2), d1, d2)

uu <- .rnorm(d, 3)
vv <- .rnorm(n, 3)
Y <- apply(uu %*% t(vv), 2, scale) + .rnorm(d, n) * 3
Y[Y < 0] <- 0

gg <- order(apply(uu,1,which.max))
kk <- apply(vv,1,which.max)

asap.data <- fileset.list(tempfile())

asap.data <- write.sparse(Matrix::Matrix(Y, sparse = T),
                          1:nrow(Y),
                          1:ncol(Y),
                          asap.data$hdr)
.info <- mmutil_info(asap.data$mtx)
```

How can we estimate cell topic proportions ASAP?

* Step 1: Create a pseudo-bulk (PB) matrix by collapsing (perhaps) "redundant" cells into one sample.

* Step 2: Perform non-negative matrix factorization (NMF) on the PB matrix.

* Step 3: Recalibrate cell-level data with a fixed dictionary matrix.

## Step 1: Fast pseudo-bulk sampling

```{r step1}
.bulk <- asap_random_bulk(asap.data$mtx,
                          asap.data$row,
                          asap.data$col,
                          asap.data$idx,
                          num_factors = 5)
```

We can squeeze `r .info$max.col` cells into `r ncol(.bulk$PB)` pseudo-bulk samples.

```{r}
Y <- stretch_matrix_columns(.bulk$PB)*100
```
Some gene-gene correlation structures are preserved in the PB data.

```{r fig_pb_y, fig.width=7, fig.height=4}
pheatmap(t(Y))
```

## Step 2: Non-negative Matrix Factorization to learn the definition of "topics"


```{r step2}
.nmf <- asap_fit_nmf(Y,
                     maxK = 5,
                     max_iter = 200,
                     svd_init = TRUE,
                     verbose = FALSE)
```

```{r}
names(.nmf)
```

```{r fig_nmf_log_x, fig.width=6, fig.height=3}
pheatmap(t(.nmf$log.beta[gg, ]), Rowv=NA, Colv=NA, scale="none", main="logBeta")
```

Some convenient routine to create the structure plot of a topic proportion matrix.

```{r}
plot.struct <- function(.prop){
    .order <- order(apply(.prop, 1, which.max))
    .melt <- melt(.prop)
    .melt$Var1 <- factor(.melt$Var1, .order)

    ggplot(.melt, aes(Var1,value,fill=as.factor(Var2))) +
        geom_bar(stat="identity") +
        scale_fill_brewer("Topics", palette = "Paired") +
        ylab("topic proportions")
}
```

```{r fig_struct_bulk, fig.width=7, fig.height=3}
.bulk.topic <- pmf2topic(.nmf$beta, .nmf$theta)
plot.struct(.bulk.topic$prop) +
    xlab("pseudobulk samples")
```

## Step 3. Cell-level recalibration to recover cell-level topic proportions


```{r step3}
.stat <- asap_topic_stat(asap.data$mtx,
                         asap.data$row,
                         asap.data$col,
                         asap.data$idx,
                         log_beta = .nmf$log.beta,
                         beta_row_names = .bulk$rownames)

R <- apply(.stat$corr, 2, scale)
```

Topic correlation statistics are already very appealing.

```{r fig_topic_cor, fig.width=7, fig.height=6}
par(mfrow=c(2,2))
for(k in 1:4){
    plot(R[,k], R[,k+1],
         col = kk+1, cex=.3,
         xlab=paste("Std. Cor.", k),
         ylab=paste("Std. Cor.", k + 1))
}
```

We can quantify topic proportions based on the correlation results.

```{r}
.pmf <- asap_topic_pmf(.nmf$beta, R, .stat$colsum, max_iter = 10)
.topic <- pmf2topic(.pmf$beta, .pmf$theta)
```

```{r fig_topic_heatmap, fig.width=8, fig.height=3}
pheatmap(t(.topic$prop[order(kk),]), Rowv=NA, Colv=NA, cluster_cols = F)
```

```{r fig_true_topic_heatmap, fig.width=8, fig.height=3}
pheatmap(t(vv[order(kk),]), Rowv=NA, Colv=NA, cluster_cols = F)
```

```{r fig_topic_pmf_cell, fig.width=7, fig.height=2}
plot.struct(.topic$prop) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    xlab(paste(nrow(R),"cells"))
```

```{r fig_topic_proj, fig.width=7, fig.height=2}
.df <- data.frame(project.proportions(.topic$prop), kk)
ggplot(.df, aes(xx,yy)) +
    theme_void() +
    facet_grid(. ~ kk) +
    geom_hex(bins=20) +
    scale_fill_distiller(direction=1)
```
